Cellular Automata and Computational Universality

### 2018-09-13 paper/paper.tex
Finished intro to universality

    As an example of a Turing machine in action, suppose we want to construct a
    one to accept the language $L = \{w\#w | w \in \{0,1\}\ast\}$.  This is not
    Context Free \cite{nonCFL}, hinting at the power behind Turing Machines with
    regard to computability. We construct the machine informally as follows:

@@

    The $\braket{M,w}$ notation is used to denote the encoding of the Turing
    Machine $M$ and its input $w$ in a way that some Turing Machine would be
    able to understand and use as input. In essence, Universal Turing Machines
    are those which are capable of simulating any other Turing Machine just from
    its description and an input, and it is from this angle that we will look at
    the universality of certain cellular automata.

@@

    \subsection{Computability and Universality of Cellular Automata}
    A Turing Machine may compute a function by starting with the input to the
    function on the tape and halting with the output of the function on the
    tape. A function $f : \Sigma \ast \rightarrow \Sigma \ast$ is computable if
    some Turing Machine $M$ halts, for every input $w$, with just the result
    $f(w)$ on the tape \cite{Computable}. Thus, any model we can devise which
    could simulate the run of a Turing Machine on some input would be capable of
    computing all the functions that the Turing Machine could, and hence such a
    model would be Turing complete.

\section{Examples}
    \subsection{Von Neumann's automaton}
    \subsection{Game of Life}
    \subsection{Fractran}
    \subsection{Langton's Ant}
    \subsection{WireWorld}
\section{Applications}
    \subsection{Cryptography}
\section{Firing Squad Synchronisation Problem}
\section{Majority Problem}

>>>>

    As an example of a Turing machine in action, suppose we want to construct a
    one to accept the language $L = \{w\#w | w \in \{0,1\}\ast\}$.  This is not
    Context Free \cite{nonCFL}, meaning Turing Machines are at least more
    powerful than Push Down Automata. We construct the machine informally as
    follows:

@@

    The $\braket{M,w}$ notation is used to denote the encoding of the Turing
    Machine $M$ and its input $w$ in a way that some Turing Machine would be
    able to understand and use as input. In essence, Universal Turing Machines
    are those which are capable of simulating any other Turing Machine just from
    its description and an input, and it is from this angle that we will look at
    the universality of certain cellular automata. In case this seems like a bit
    of a leap, we can simply describe the Universal Turing Machine $U$ as
    follows:

    \begin{turing}{U}{on input $\braket{M,w}$, where $M$ is a Turing Machine and
        $w$ is a string:}
        \item Run $M$ on input $w$
        \item If $M$ ever enters its accept state, accept; if $M$ ever enters
            its reject state, reject
    \end{turing}

    Note that $U$ will never halt if $M$ never halts, and so while it recognises
    the language $A_{TM}$, it does not decide it. In fact, $A_{TM}$ is called
    the Halting Problem and is undecidable: there is no way for an algorithm to
    determine whether some Turing Machine $M$ may or may not halt on input $w$,
    without simulating it itself.

@@

    \subsection{Universality of Game of Life}
    A Turing Machine may compute a function by starting with the input to the
    function on the tape and halting with the output of the function on the
    tape. A function $f : \Sigma \ast \rightarrow \Sigma \ast$ is computable if
    some Turing Machine $M$ halts, for every input $w$, with just the result
    $f(w)$ on the tape \cite{Computable}. Thus, any model we can devise which
    could simulate the run of a Turing Machine on some input would be capable of
    computing all the functions that the Turing Machine could, and hence such a
    model would be Turing complete.

    The key to Game of Life's universal computation is its ability to create
    infinite growth, for example by using glider guns. Guns can then be
    positioned to shoot at blocks in specific locations, and doing so correctly
    can be used to move the block closer or further away. This sliding block
    memory can be used to simulate a counter \cite{Counter}. Logical AND, OR,
    and NOT gates can also be constructed using gliders, with the presence of a
    glider signifying a 1, and 0 otherwise. These three Boolean operators are
    theoretically necessary and sufficient to build a Universal Turing Machine
    \cite{BoolTM}. A fully functioning Turing Machine was designed to completion
    by Paul Rendell on 2\textsuperscript{nd} April, 2000 \cite{golTM}, which
    doubled the number on the input tape. Further extensions led to the
    development of Universal Turing Machines in February 2010 and March 2011,
    both of which were capable of simulating the Turing Machine from 2000
    \cite{golUTM1,golUTM2}.

    \begin{figure}[h]
        \includegraphics{}
        \caption{}
        \label{}
    \end{figure}

\section{More cellular automata}
    \subsection{WireWorld}
    \subsubsection{Computational Universality}
    \subsection{Rule 110}
    \subsubsection{Computational Universality}
    \subsection{Von Neumann's automaton}
    \subsection{Fractran}
    \subsection{Langton's Ant}
\section{Applications}
    \subsection{Cryptography}
\section{Firing Squad Synchronisation Problem}
\section{Majority Problem}
